Setting up Data

### 2020-06-02 doc/MS_v2.Rmd
updating the Rdata files, reduce redundancyes

new - Images were capture using a Zeiss Axioplan 2 microscope with AxioLab camera and AxioVision software (Zeiss, Cambridge, UK). Preprocessing, including cropping, noise reduction, and histogram adjustments, was performed using Photoshop (v13.0). Image file names were anonymized before manual scoring of MLH1 or DMC1 foci using Photoshop.

@@

*new-* To estimate the number of crossovers across the genome, we counted MLH1 foci. MLH1 foci were counted in cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes; 20 acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality score ranging from 1 (best) to 5 (worst) was assigned to each cell based on visual appearance of staining and spread of bivalents. Cells with a score of 5 were excluded from the final analysis. Distributions of MLH1 count per cell were visually inspected for normality (Supplemental figure 1). MLH1 foci located on the XY in spermatocytes were excluded from counts. In addition to MLH1 counts, we measured several traits to further characterize the recombination landscape. To estimate the number of double-strand breaks, a minority of which lead to crossovers, mean DMC1 foci per cell was quantified for a single male from a subset of strains ( _musculus^PWD^_, _musculus^MSM^_, _domesticus^WSB^_, and _domesticus^G^_ ). SC morphology and CREST foci number was used to stage spermatocytes as early zygotene or late zygotene.

To measure bivalent SC length, two image analysis algorithms were used. The first algorithm estimates the total (summed) SC length across bivalents for individual cells (@wang2019_SC). The second algorithm estimates the SC length of individual bivalents (@peterson2019). Both algorithms apply a ‘skeletonizing’ transformation to synapsed chromosomes that produces a single, pixel-wide ‘trace’ of the bivalent shape.

Total SC length per cell was quantified from pachytene cell images [@wang2019_SC]. To reduce algorithmic errors in SC isolation, outliers were visually identified at the mouse level and removed from the data set. Mouse averages were calculated from cell-wide total SC lengths in `r formatC( length(  (Total.SC_DF  %>% filter(species == "M.musculus")  )$Random.Name), format="d", big.mark=",")` -- 3,371 out of `r formatC( length(  (MLH1_data %>% filter(species == "M.musculus") )$Random.Name), format="d", big.mark=",")`-- 4,143 cells with MLH1 counts.

@@

SC length of individual bivalents was quantified in pachytene cell images [@peterson2019]. The DNA CrossOver algorithm [@peterson2019] isolates single, straightened bivalent shapes, returning SC length, location of MLH1 foci, and location of CREST (centromere) foci. The algorithm substantially speeds the accurate measurement of bivalents, but it sometimes interprets overlapping bivalents as single bivalents. In our data set, average proportions of bivalents per cell isolated by the algorithm ranged from 
`r min(algorithm.stat.category$mean_pass.rate)` -- 0.51 ( _molossinus^MSM^_ male) to`r max(algorithm.stat.category$mean_pass.rate)` -- 0.72 ( _musculus^KAZ^_ female). From the total set of cell images, `r formatC( length(unique(cleanCurated_BivData %>% filter(SC.pass == "1"))$fileName ), format="d", big.mark=",")` -- 10,458 bivalent objects were isolated by the algorithm. Following a manual curation, `r formatC( length(unique(cleanCurated_BivData %>% filter(hand.foci.count != ""))$Obj.ID ), format="d", big.mark=",")` -- 9,829 single-bivalent observations remained. The accuracy of the algorithm is high compared to hand measures after this curation step [@peterson2019]. The curated single bivalent data supplements our cell wide MLH1 count data with MLH1 foci counts for single bivalents. The proportions of bivalents with the same number of MLH1 foci where compared across strains by a chi-squared test using prop.test().

To account for confounding effects of sex chromosomes from pooled samples of bivalents, we also considered a reduced data set including only bivalents with SC lengths below the 2nd quartile in cells with at least 17 of 20 single bivalent measures. This “short bivalent” data set included the four or five shortest bivalents, and excluded the -- X bivalent in oocytes. A total of `r sum(Short.biv_mouse_table$nobs )` -- 678 short bivalents were isolated from `r sum(  (Short.biv_mouse_table  %>% filter(sex=="female") )$ncells)` -- 103 oocytes and `r sum(  (Short.biv_mouse_table  %>% filter(sex=="male") )$ncells)` -- 37 spermatocytes. Although this smaller data set has decreased power, it offers a more comparable set of single bivalents to compare between the sexes. A “long bivalent” data set was formed from those bivalents above the 4th quartile in SC lengths per cell. `r sum(Long.biv_mouse_table$n_obs)`  number of long bivalents were isolated from `r sum(  (Long.biv_mouse_table  %>% filter(sex=="female") )$ncells)` oocytes and xx `r sum(  (Long.biv_mouse_table  %>% filter(sex=="male") )$ncells)` spermatocytes.

@@

We used a series of statistical models to interpret patterns of variation in the recombination traits we measured (Table 2). We first constructed a mixed model (M1) using the lmer() from the lmer4 package (cite) in R (v3.5.2) [@Rstudio]. In this model, strain was coded as a random effect, with significance evaluated using a likelihood ratio test (using exactRLRT()). Subspecies, strain, and their interaction were coded as fixed effects,  with significance evaluated using a chi-square test comparing the full and reduced models (drop1() and anova()). As observations were at the level of single mouse (mouse average) was uniquely coded within the dataset, nesting was implicit, each strain only occurs within one strain, and was not explicitly coded in our linear and mixed models.

We used the subspecies effect to quantify divergence between subspecies and the (random) strain effect to quantify variation within subspecies in a sex-specific manner. In separate analyses, we considered mouse averages as dependent variables for each of the following traits: MLH1 count per cell, variance in MLH1 count per cell, coefficient of variation in MLH1 count per cell, total SC length per cell, single bivalent SC length per cell, IFD~raw~, IFD~norm~, and average MLH1 position (for single-focus bivalents). Four additional models containing only fixed effects (M2-M5) (Table 2) were used to further investigate results obtained from the initial mixed model.

@@

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` -- 14 wild-derived inbred strains sampled from three subspecies of house mice ( _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_ ) and two additional species of Mus ( _M. spretus_ and _M. spicilegus_ ). Mean MLH1 foci counts for `r sum(main.table$Nmice)` -- 161 mice were quantified from an average of `r mean( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T)` -- 22.0 spermatocytes per male (for a total of `r sum( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T)` -- 1,867 spermatocytes) and `r mean( (Mouse.Table_HQ %>% filter(sex == "female") )$Ncells, na.rm = T)` -- 18.5 oocytes per female (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells), format="d", big.mark=",")` -- 1,409 oocytes).
Graphical comparisons reveal sex-specific dynamics to the evolution of genome-wide recombination rate (Figure 1A). First, MLH1 focus counts differ between females and males in most strains. Second, the relationship between MLH1 foci in the sexes varies among strains. Although most strains show more MLH1 foci in females, three strains - _musculus^PWD^_, _musculus^SKIVE^_, and _molossinus^MSM^_ - exhibit higher MLH1 focus counts in males. In females, numbers of MLH1 foci are evenly distributed around the sex-wide mean of approximately 25 (Figure 1B). In stark contrast, males largely separate into two groups of strains with high numbers (near 30) and low numbers (near 23 MLH1 foci per cell) of foci (Figure 1C).

To further situate variation in house mouse recombination rate within an evolutionary framework, we fit a series of statistical models including subspecies, strain, and sex, to mean MLH1 focus counts from `r length(DF.HetC.MixedModel.HQ$mean_co)` -- 187 mice. **to best investigate sex, will limited our analysis to strains with sufficient observations from both sexes (Table 1).**

We began with a full mixed model (M1; see Materials and Methods), showing that strain (random effect p < 10^{-6} -- p < `r 1/10000`), sex (p = 1.5510^{-8} -- `r MLH1.MMlmer_results$sex.results`), subspecies (p=1.7210^{-4} -- `r MLH1.MMlmer_results$sub.sp`), and subspecies*sex (p = 3.110^{-5} -- `r MLH1.MMlmer_results$interaction.pval` ) each significantly affect MLH1 focus count. After a model including all factors as fixed effects (M2) revealed only weak contributions of subspecies, we focused on additional models designed to illuminate the role of strain and sex. A model with only these two variables (M3) identified two strains with particularly strong effects on MLH1 focus count:  _musculus^MSM^_ (p = 3.9910^{-6} -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]`) and _domesticus^G^_  (p = 1.0410^{-6} -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` ). In addition, two strains exhibit strain-by-sex interactions: _molossinus^MSM^_  (p = 1.2610^{-4} -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]` ) and _musculus^PWD^_ (p = 3.8610^{-4} -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]` ).

We next fit models separately for  `r length(male_mouse.avs_for.analysis$mouse)` -- 192 males and `r length(female_mouse.avs_for.analysis$mouse)` -- 144 females (M4). In the male dataset, three strains affect MLH1 focus count (as observed in Figure 1C): _musculus^PWD^_ (p = 6.3110^{-8} -- `r sum_MLH1.male.glm.M4$coefficients[,4][8]` ; effect = `r sum_MLH1.male.glm.M4$coefficients[8]` -- 6.1 foci), and _musculus^SKIVE^_ (p = 0--`r sum_MLH1.male.glm.M4$coefficients[,4][10]` ; effect = 3.8-- `r sum_MLH1.male.glm.M4$coefficients[10]` ), and _molossinus^MSM^_  (p=2.4210^{-12} -- `r sum_MLH1.male.glm.M4$coefficients[,4][9]`; effect = `r sum_MLH1.male.glm.M4$coefficients[9]` -- 7.0).

@@

Our survey of recombination rate across house mice provides an opportunity to determine whether sex differences in chromatin compaction (SC length) are reversed when heterochiasmy is reversed. In all strains except _musculus^SKIVE^_, females have longer SCs than males, whether SC length was estimated as the total length across bivalents or as the length of short bivalents (all t-test below; p < 0.05; except short bivalents for SKIVE p = `r Q1_short.SC_t.test.SKIVE$p.value` ).

In the dataset of short bivalents (to which the female X bivalent does not contribute), female to male ratios of mouse mean SC length range from 1.15 -- `r Q1SC.mus.av.short.biv_DIFF_PWD` ( _musculus^PWD^_) to 1.49 -- `r Q1SC.mus.av.short.biv_DIFF_WSB` ( _domesticus^WSB^_) across strains.

That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC  (M1: p=`r Q1.lmer_totSC_M1_results$sex.results`; M2: p=`r sum.Q1.glm_totSC_M2$coefficients[,c(1,4)][9,2]`; M3: p =`r sum.Q1.glm_totSC_M3$coefficients[,c(1,4)][8,2]`) and short bivalent measures (M1: p= `r Q1.lmer_short_M1_results$sex.results`; M2: p= `r summ.Q1.glm_short.SC_M2$coefficients[,c(1,4)][8,2]`; M3: `r summ.Q1.glm_short.SC_M3$coefficients[,c(1,4)][8,2]`).

The existence of some subspecies and strain effects (M1 interaction: `r Q1.lmer_short_M1_results$interaction.pval`; M2: strain-SKIVE `r summ.Q1.glm_short.SC_M2$coefficients[,c(1,4)][7,]`,  SKIVE * male `r summ.Q1.glm_short.SC_M2$coefficients[,c(1,4)][13,]`;  M3: SKIVE * male  `r summ.Q1.glm_short.SC_M3$coefficients[,c(1,4)][13,]`) further indicates that SC length has evolved among strains and among subspecies.

In summary, two approaches for measuring SC length demonstrate that females have longer SCs (chromosome axes), even in strains in which males have more MLH1 foci per cell. This pattern further suggests that in high-recombination strains, spermatocytes have less space in which to position additional crossovers than do oocytes.

@@

```{r Q1.F1.pos.LONG.SHORT.BIV_scatter, echo=FALSE}
#real.long.bivData
Q1.short.biv_scatr_box   <- ggplot( (real.short.bivData %>% filter(hand.foci.count == 1) ), aes(x=sex, y=PER_Foci_1, color=strain) )+geom_jitter( )+ geom_boxplot(aes(y= PER_Foci_1 ),alpha=.3 ) + scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales = "free")+ylim(c(0,1.1))+ggtitle("Mouse av. \nNormalized 1CO Foci pos \nShort Biv")

Q1.long.biv_scatr_box   <- ggplot((real.long.bivData %>% filter(hand.foci.count == 1) ), aes(x=sex, y=PER_Foci_1, color=strain) )+geom_jitter( )+ geom_boxplot(aes(y= PER_Foci_1 ),alpha=.3 ) + scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales = "free")+ylim(c(0,1.1))+ggtitle("Mouse av. \nNormalized 1CO Foci pos \nLong Biv")

@@

Our conclusions are accompanied by several caveats. First, MLH1 foci only identify interfering crossovers **(REF)**. Although most crossovers (XX%; REF) belong to this class, our approach likely underestimated genome-wide recombination rates. Evolution of the number of non-interfering crossovers is a topic worth examining. A second limitation is that our investigation of crossover positions was confined to the relatively low resolution possible with immunofluorescent cytology. Positioning crossovers with higher resolution could reveal additional evolutionary patterns. Finally, the panel of inbred lines we surveyed may not be representative of recombination rate variation within and between subspecies of house mice. We considered most available wild-derived inbred lines, but house mice have a broad geographic distribution. Nevertheless, we expect our conclusion that recombination evolves in a sex-specific manner to be robust to geographic sampling because differences between females and males exist for the same set of inbred strains.

>>>>

Images were capture using a Zeiss Axioplan 2 microscope with AxioLab camera and AxioVision software (Zeiss, Cambridge, UK). Preprocessing, including cropping, noise reduction, and histogram adjustments, was performed using Photoshop (v13.0). Image file names were anonymized before manual scoring of MLH1 or DMC1 foci using Photoshop.

@@

To estimate the number of crossovers across the genome, we counted MLH1 foci. MLH1 foci were counted in cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes; 20 acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality score ranging from 1 (best) to 5 (worst) was assigned to each cell based on visual appearance of staining and spread of bivalents. Cells with a score of 5 were excluded from the final analysis. Distributions of MLH1 count per cell were visually inspected for normality (Supplemental figure 1). MLH1 foci located on the XY in spermatocytes were excluded from counts. In addition to MLH1 counts, we measured several traits to further characterize the recombination landscape. To estimate the number of double-strand breaks, a minority of which lead to crossovers, mean DMC1 foci per cell was quantified for a single male from a subset of strains ( _musculus^PWD^_, _musculus^MSM^_, _domesticus^WSB^_, and _domesticus^G^_ ). SC morphology and CREST foci number was used to stage spermatocytes as early zygotene or late zygotene.

To measure bivalent SC length, two image analysis algorithms were used. The first algorithm estimates the total (summed) SC length across bivalents for individual cells (@wang2019_SC). The second algorithm estimates the SC length of individual bivalents (@peterson2019). Both algorithms apply a ‘skeletonizing’ transformation to synapsed chromosomes that produces a single, pixel-wide ‘trace’ of the bivalent shape. Total SC length per cell was quantified from pachytene cell images [@wang2019_SC]. To reduce algorithmic errors in SC isolation, outliers were visually identified at the mouse level and removed from the data set. Mouse averages were calculated from cell-wide total SC lengths in `r formatC( length(  (Total.SC_DF  %>% filter(species == "M.musculus")  )$Random.Name), format="d", big.mark=",")` -- 3,371 out of `r formatC( length(  (MLH1_data %>% filter(species == "M.musculus") )$Random.Name), format="d", big.mark=",")`-- 4,143 cells with MLH1 counts.

@@

SC length of individual bivalents was quantified in pachytene cell images [@peterson2019]. The DNA CrossOver algorithm [@peterson2019] isolates single, straightened bivalent shapes, returning SC length, location of MLH1 foci, and location of CREST (centromere) foci. The algorithm substantially speeds the accurate measurement of bivalents, but it sometimes interprets overlapping bivalents as single bivalents. In our data set, average proportions of bivalents per cell isolated by the algorithm ranged from 
`r min(algorithm.stat.category$mean_pass.rate)` -- 0.51 ( _molossinus^MSM^_ male) to`r max(algorithm.stat.category$mean_pass.rate)` -- 0.72 ( _musculus^KAZ^_ female). From the total set of pachytene cell images, `r formatC(  length(unique(Curated_BivData$Obj.ID ) ) , format="d", big.mark=",")` -- 10,458 bivalent objects were isolated by the algorithm. Following a manual curation, `r formatC( length(unique(cleanCurated_BivData %>% filter(hand.foci.count != ""))$Obj.ID ), format="d", big.mark=",")` -- 9,829 single-bivalent observations remained. The accuracy of the algorithm is high compared to hand measures after this curation step [@peterson2019]. The curated single bivalent data supplements our cell wide MLH1 count data with MLH1 foci counts for single bivalents. The proportions of bivalents with the same number of MLH1 foci where compared across strains by a chi-squared test using prop.test().

To account for confounding effects of sex chromosomes from pooled samples of bivalents, we also considered a reduced data set including only bivalents with SC lengths below the 2nd quartile in cells with at least 17 of 20 single bivalent measures. This “short bivalent” data set included the four or five shortest bivalents, and excluded the -- X bivalent in oocytes. A total of `r sum(Short.biv_mouse_table$nobs )` -- 678 short bivalents were isolated from `r sum(  (Short.biv_mouse_table  %>% filter(sex=="female") )$ncells)` -- 103 oocytes and `r sum(  (Short.biv_mouse_table  %>% filter(sex=="male") )$ncells)` -- 37 spermatocytes. Although this smaller data set has decreased power, it offers a more comparable set of single bivalents to compare between the sexes. A “long bivalent” data set was formed from those bivalents above the 4th quartile in SC lengths per cell. A total of `r sum(Long.biv_mouse_table$n_obs)`  long bivalents were isolated from `r sum(  (Long.biv_mouse_table  %>% filter(sex=="female") )$Ncells)` oocytes and `r sum(  (Long.biv_mouse_table  %>% filter(sex=="male") )$Ncells)` spermatocytes.

@@

We used a series of statistical models to interpret patterns of variation in the recombination traits we measured (Table 2). We first constructed a mixed model (M1) using the lmer() from the lmer4 package [@lme4] in R (v3.5.2) [@Rstudio]. In this model, strain was coded as a random effect, with significance evaluated using a likelihood ratio test (using exactRLRT() from RLRsim [@RLRsim]). Subspecies, strain, and their interaction were coded as fixed effects,  with significance evaluated using a chi-square test comparing the full and reduced models (drop1() and anova()) [@lme4]. As observations were at the level of single mouse (mouse average) was uniquely coded within the dataset, nesting was implicit, each strain only occurs within one strain, and was not explicitly coded in our linear and mixed models.

We used the subspecies effect to quantify divergence between subspecies and the (random) strain effect to quantify variation within subspecies in a sex-specific manner. In separate analyses, we considered mouse averages as dependent variables for each of the following traits: MLH1 count per cell, total SC length per cell, single bivalent SC length per cell, IFD~raw~, IFD~norm~, and average MLH1 position (for single-focus bivalents). Four additional models containing only fixed effects (M2-M5) (Table 2) were used to further investigate results obtained from the initial mixed model.

@@

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` -- 14 wild-derived inbred strains sampled from three subspecies of house mice ( _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_ ) and two additional species of Mus ( _M. spretus_ and _M. spicilegus_ ). Mean MLH1 foci counts for `r sum(main.table$Nmice)` -- 161 mice were quantified from an average of `r mean( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T)` -- 22.0 spermatocytes per male (for a total of `r formatC( sum( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T), format="d", big.mark=",")` -- 1,867 spermatocytes) and `r mean( (Mouse.Table_HQ %>% filter(sex == "female") )$Ncells, na.rm = T)` -- 18.5 oocytes per female (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells), format="d", big.mark=",")` -- 1,409 oocytes).
Graphical comparisons reveal sex-specific dynamics to the evolution of genome-wide recombination rate (Figure 1A). First, MLH1 focus counts differ between females and males in most strains. Second, the relationship between MLH1 foci in the sexes varies among strains. Although most strains show more MLH1 foci in females, two strains - _musculus^PWD^_ and _molossinus^MSM^_ - exhibit higher MLH1 focus counts in males. While _musculus^SKIVE^_ has a very low degree of sex differences in MLH1 focus counts. In females, numbers of MLH1 foci are evenly distributed around the sex-wide mean of approximately 25 (Figure 1B). In stark contrast, males largely separate into two groups of strains with high numbers (near 30) and low numbers (near 23 MLH1 foci per cell) of foci (Figure 1C).

To further situate variation in house mouse recombination rate within an evolutionary framework, we fit a series of statistical models including subspecies, strain, and sex, to mean MLH1 focus counts from `r length( DF.HetC.MixedModel.HQ$mean_co)` -- 187 mice. To investigate the effects of sex, we limited our analysis to `r length( unique( DF.HetC.MixedModel.HQ$strain) )` strains with sufficient observations from both sexes (Table 1).

We began with a full mixed model (M1; see Materials and Methods), showing that strain (random effect p < 10^{-6} -- p < `r 1/10000`), sex (p = 1.5510^{-8} -- `r MLH1.MMlmer_results$sex.results`), subspecies (p=1.7210^{-4} -- `r MLH1.MMlmer_results$sub.sp`), and subspecies*sex (p = 3.110^{-5} -- `r MLH1.MMlmer_results$interaction.pval` ) each significantly affect MLH1 focus count. After a model including all factors as fixed effects (M2) revealed only weak contributions of subspecies, we focused on additional models designed to illuminate the role of strain and sex. A model with only these two variables (M3) identified two strains with particularly strong effects on MLH1 focus count:  _musculus^MSM^_ (p = 3.9910^{-6} -- p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]`) and _domesticus^G^_  (p = 1.0410^{-6} -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` ). In addition, two strains exhibit strain-by-sex interactions: _molossinus^MSM^_  (p = 1.2610^{-4} -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]` ) and _musculus^PWD^_ (p = 3.8610^{-4} -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]` ).

We next fit models separately for  `r length(male_mouse.avs_for.analysis$mouse)` -- 192 males and `r length(female_mouse.avs_for.analysis$mouse)` -- 144 females (M4). In the male dataset, three strains affect MLH1 focus count (as observed in Figure 1C): _musculus^PWD^_ (p = 6.3110^{-8} -- `r sum_MLH1.male.glm.M4$coefficients[,4][8]` ; effect = `r sum_MLH1.male.glm.M4$coefficients[8]` -- 6.1 foci), and _musculus^SKIVE^_ (p = `r sum_MLH1.male.glm.M4$coefficients[,4][10]` ; effect = 3.8-- `r sum_MLH1.male.glm.M4$coefficients[10]` ), and _molossinus^MSM^_  (p=2.4210^{-12} -- `r sum_MLH1.male.glm.M4$coefficients[,4][9]`; effect = `r sum_MLH1.male.glm.M4$coefficients[9]` -- 7.0).

@@

Our survey of recombination rate across house mice provides an opportunity to determine whether sex differences in chromatin compaction (SC length) are reversed when heterochiasmy is reversed. In all strains except _musculus^SKIVE^_, females have longer SCs than males, whether SC length was estimated as the total length across bivalents or as the length of short bivalents (all t-test below; p < 0.05; except short bivalents for SKIVE p = `r Q1_short.SC_t.test.SKIVE$p.value` ). In the reduced dataset of short bivalents (to which the female X bivalent does not contribute), female to male ratios of mouse mean SC length range from  `r Q1SC.mus.av.short.biv_DIFF_PWD` ( _musculus^PWD^_) to `r Q1SC.mus.av.short.biv_DIFF_WSB` ( _domesticus^WSB^_) across strains.

That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC  (M1: p=`r Q1.lmer_totSC_M1_results$sex.results`; M2: p=`r sum.Q1.glm_totSC_M2$coefficients[,c(1,4)][9,2]`; M3: p =`r sum.Q1.glm_totSC_M3$coefficients[,c(1,4)][8,2]`) and short bivalent measures (M1: p= `r Q1.lmer_short_M1_results$sex.results`; M2: p= `r summ.Q1.glm_short.SC_M2$coefficients[,c(1,4)][8,2] / 10000`; M3: p = `r summ.Q1.glm_short.SC_M3$coefficients[,c(1,4)][8,2] / 10000`). The existence of some subspecies and strain effects (M1 interaction: p = `r Q1.lmer_short_M1_results$interaction.pval`; M2: strain-SKIVE `r summ.Q1.glm_short.SC_M2$coefficients[7,c(1,4)][2]`,  SKIVE * male p= `r summ.Q1.glm_short.SC_M2$coefficients[13,c(1,4)][2]`) further indicates that SC length has evolved among strains and among subspecies. In summary, two approaches for measuring SC length demonstrate that females have longer SCs (chromosome axes), even in strains in which males have more MLH1 foci per cell. This pattern further suggests that in high-recombination strains, spermatocytes have less space in which to position additional crossovers than do oocytes.

@@

@@

Our conclusions are accompanied by several caveats. First, MLH1 foci only identify interfering crossovers [@holloway2008mus81]. Although most crossovers (XX%; REF) belong to this class, our approach likely underestimated genome-wide recombination rates. Evolution of the number of non-interfering crossovers is a topic worth examining. A second limitation is that our investigation of crossover positions was confined to the relatively low resolution possible with immunofluorescent cytology. Positioning crossovers with higher resolution could reveal additional evolutionary patterns. Finally, the panel of inbred lines we surveyed may not be representative of recombination rate variation within and between subspecies of house mice. We considered most available wild-derived inbred lines, but house mice have a broad geographic distribution. Nevertheless, we expect our conclusion that recombination evolves in a sex-specific manner to be robust to geographic sampling because differences between females and males exist for the same set of inbred strains.

### 2020-06-02 src/Result_Setup.Rmd
updating the Rdata files, reduce redundancyes

#what was this for
#all.cell.MM.HQ <- ddply(.data=DF.HetC.MixedModel.HQ,
#                 .(sex, strain),
#                 summarize,
#                 all.cells = sum(Ncells)
#)
```

@@

```{r table.long.biv, echo=FALSE}

@@

#2) Long Biv for male.poly
#just males, just Musculus
Long.biv_mouse_table_male.poly <- Long.biv_mouse_table

Long.biv_mouse_table_male.poly <- Long.biv_mouse_table_male.poly[Long.biv_mouse_table_male.poly$sex == "male",]

#table(droplevels( Long.biv_mouse_table_male.poly$category) ) #for double checking
```

```{r table.short.biv, echo=FALSE}

#real.short.bivData

@@

#unordered for the mixed models (move this to the MM blocks...?)
Short.biv_mouse_table$subsp <- factor(Long.biv_mouse_table$subsp, ordered = FALSE)
Short.biv_mouse_table$sex <- factor(Long.biv_mouse_table$sex, ordered = FALSE)
Short.biv_mouse_table$strain <- factor(Long.biv_mouse_table$strain, ordered = FALSE)

@@

#1) Long Biv for 
#remove all the non-used strains!!
#Short.biv_mouse_table_HetC <- Short.biv_mouse_table

#Short.biv_mouse_table_HetC <- Short.biv_mouse_table_HetC[Long.biv_mouse_table$strain != "AST",]

#Short.biv_mouse_table_HetC <- Short.biv_mouse_table_HetC[Curated_BivData$strain != "TOM",]

#Short.biv_mouse_table_HetC <- Short.biv_mouse_table_HetC[Long.biv_mouse_table_HetC$strain != "CZECH",]

#Long.biv_mouse_table_HetC$strain <- droplevels(Long.biv_mouse_table_HetC$strain)

#remove na's
#Long.biv_mouse_table_HetC <- #Long.biv_mouse_table_HetC[!(is.na(Long.biv_mouse_table_HetC$mouse) | #Long.biv_mouse_table_HetC$mouse==""), ]

```{r Q1.Short.Long_biv.setup, echo=FALSE, eval=FALSE}
#moved this to setupdata

#REAL.long_mouse_table
#REAL.short_mouse_table
#Long.biv_mouse_table

#make the Q1 strain list
Q1.real.short_mouse_table <- Long.biv_mouse_table %>% filter( (strain != "CZECH") & (strain != "MOLF") )

Q1.real.long_mouse_table <- Long.biv_mouse_table %>% filter( (strain != "CZECH") & (strain != "MOLF") )

@@

```{r Q2_tot.sc_plot, echo=FALSE, warning=FALSE, eval=FALSE}

#totSC.mouse.av.table <- totSC.mouse.av.table %>% filter(strain!="CAST" & strain !="F1"& strain !="SPIC" & strain !=  "CAROLI" & strain !="SPRET"& strain !="HMI"& strain !="PERC")

#MLH1_data <- MLH1_data %>%
#  arrange(strain, sex, mouse) %>%
#  mutate(file.name = factor(file.name))

mouse.av_totsc_male <- ggplot(data= totSC.mouse.av.table %>% filter(sex == "male"), aes(y = mean.skel, x=mouse, color=strain))+ geom_point()+scale_color_manual(values=colors_of_strains)+
  facet_wrap(Rec.group~subsp, scales="free")+ theme(axis.title.x=element_blank(),        axis.text.x=element_blank(), axis.ticks.x=element_blank()) + ylim(c(1200,1850 ))+ggtitle("Mouse av. total SC area")

@@

```
The molf male t-tests don't work because I only have total Sc measures for 1 MOLF male. to of the todo list if I get back to lab.

@@

@@

@@

@@

@@

@@

#mouse av for long biv
Male.cell.av.long.biv.table <- ddply(male.long.biv, c("fileName"), summarise,

                        n_obs = length(Obj.ID),
                         mean.SC = mean(chromosomeLength)
                         )
Male.cell.av.long.biv.table <- add_mouse(Male.cell.av.long.biv.table)
Male.cell.av.long.biv.table <- add_strain(Male.cell.av.long.biv.table)
Male.cell.av.long.biv.table <- add_subsp(Male.cell.av.long.biv.table)

Male.cell.av.long.biv.table$strain <- factor(Male.cell.av.long.biv.table$strain, ordered = FALSE)
Male.cell.av.long.biv.table$strain <- factor(Male.cell.av.long.biv.table$strain, ordered = FALSE)
Male.cell.av.long.biv.table$strain <- factor(Male.cell.av.long.biv.table$strain, ordered = FALSE)

Q2mouse.av.long.biv.table$Rec.group.exclu <- ifelse(grepl("PWD", Q2mouse.av.long.biv.table$strain), 1, 
                                ifelse(grepl("MSM", Q2mouse.av.long.biv.table$strain), 1, 0))

Q2mouse.av.short.biv.table$Rec.group.exclu <- ifelse(grepl("PWD", Q2mouse.av.short.biv.table$strain), 1, 
                                ifelse(grepl("MSM", Q2mouse.av.short.biv.table$strain), 1, 0))
```

@@

mouse.shoo.long.biv <- ggplot( Long.biv_mouse_table, aes(y=mean.sc, x=strain, color=strain))+geom_jitter()+facet_wrap(~subsp, scales = "free")+scale_color_manual(values=colors_of_strains)+ylim(c(75,140))+geom_boxplot(alpha=.2) + ggtitle("Mouse av. \nMale Long Biv SC Lengths")

```{r Q2.glm.SC.long, echo=FALSE}

@@

@@

```{r Q2.IFD.glm.test, echo=FALSE, warning=FALSE, eval=FALSE}

>>>>

#remove male only data
remove_strains <- c("TOM","AST","CZECH","CAST","HMI","SPRET","CAROLI", "F1", "PERC", "SPIC")

DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel.HQ[! DF.HetC.MixedModel.HQ$strain %in% remove_strains,]

@@

```{r table.long.biv, echo=FALSE, eval =FALSE}
#This table is already made in setupData

@@

```{r table.short.biv, echo=FALSE, eval=FALSE}

#this table is made in SetUpData
#real.short.bivData

@@

#unordered for the mixed models (move this to the MM blocks...?)
Short.biv_mouse_table$subsp <- factor(Short.biv_mouse_table$subsp, ordered = FALSE)
Short.biv_mouse_table$sex <- factor(Short.biv_mouse_table$sex, ordered = FALSE)
Short.biv_mouse_table$strain <- factor(Short.biv_mouse_table$strain, ordered = FALSE)

@@

@@

@@

The molf male t-tests don't work because I only have total Sc measures for 1 MOLF male. to of the

```{r Q2.totSC.ttest_old, echo=FALSE, eval=FALSE}

todo list if I get back to lab.

@@

```{r Q2.short.long.ttest.old, echo=FALSE, eval=FALSE}

@@

```{r Q2.SC_HvL_allBiv.old, echo=FALSE, eval=FALSE}

@@

```{r Q2.SC_across.subsp_short.long_biv.old, echo=FALSE, eval=FALSE}

@@

```{r Q2.glm.SC.long.old, echo=FALSE, eval=FALSE}

@@

```{r Q2.glm.SC.short.old, echo=FALSE, eval=FALSE}
#Q2mouse.av.short.biv.table
#mouse.av.short.biv.table

@@

@@

@@

```{r Q2.IFD.glm.test.old, echo=FALSE, warning=FALSE, eval=FALSE}

@@

### 2020-06-02 src/SetupData.Rmd
updating the Rdata files, reduce redundancyes

                        Nmice = length(unique(mouse)),
                        Ncells  = length(Obj.ID),

@@

                        Nmice = length(unique(mouse)),
                        Ncells  = length(Obj.ID),

>>>>

                        Nmice = length(unique(mouse)),
                        Ncells  = length( unique(fileName)),
                      n_obs  = length(Obj.ID),
                      mean_SC = as.numeric(format(round(  mean(chromosomeLength), 3 ), nsmall=3) ),
                        cV = cv(chromosomeLength),
                        var = as.numeric(format(round(   var(chromosomeLength),3), nsmall=3) ),
                        sd   = round(sd(chromosomeLength), 3),
                        se   = round(sd / sqrt(Ncells), 3),

@@

                        Nmice = length(unique(mouse)),
                        Ncells  = length( unique(fileName) ),
                        n_obs  = length(Obj.ID),
